---
title: "Does the mouse 39 °C Treg stress signature appear in JIA synovial-fluid T cells?"
subtitle: "Mouse WT_heat (39 °C Treg) signature scored in JIA synovial-fluid vs paired-blood sorted T cells (GSE160097)"
format:
  html:
    toc: true
    toc-location: left
    df-print: paged
    embed-resources: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.align = "center", dpi = 150, out.width = "95%")
.find_root <- function(sentinel = "02_analysis/config/analysis_config.yaml", start = getwd()) {
  d <- normalizePath(start)
  while (!file.exists(file.path(d, sentinel))) {
    parent <- dirname(d); if (identical(parent, d)) stop("root not found"); d <- parent
  }
  d
}
knitr::opts_knit$set(root.dir = .find_root())
```

```{r libs}
suppressPackageStartupMessages({ library(dplyr); library(knitr) })
here <- function(...) file.path(getwd(), ...)
# Reusable interactive running-sum widget (carved out of the inline chunk below so
# any chunk / any qmd can call it): defines build_runsum_widget().
source("02_analysis/helpers/runsum_widget.R")
```

*Notebook run `r format(Sys.time(), '%Y-%m-%d %H:%M UTC', tz='UTC')`.*

## The question this dataset answers

Does the mouse 39 °C Treg signature show interpretable enrichment in JIA synovial-fluid
(SF) Tregs relative to paired peripheral blood? This review settles the one-sentence claim
the result licenses — enriched, with NES/FDR, or a clean null (both are reportable) — and
whether any enrichment is Treg-preferential relative to Tcon/CD8 or instead pan-T (a weaker,
different claim). It also names which single figure carries the result. If the signal is
null, two narratives remain open: a niche account that keeps myeloid/fibroblast scoring in
view for the tissue datasets, or "the axis stays undetectable in human arthritis T cells", in
which case the SAVI dataset remains the positive control.

> A clean null is a *reported result* these same figures carry on equal footing, because the
> analysis rests on donor-level effect sizes with confidence intervals read on their own
> scale, not on a significance threshold.

## Primary evidence — `WT_heat` NES on donor-pseudobulk SF-vs-PB ranked lists

```{r load-gsea}
tdir <- here("03_results/05_scoring/tables")
tag  <- c(Treg = "treg", Tcon = "tcon", CD8 = "cd8")
gsea <- do.call(rbind, lapply(names(tag), function(pop) {
  f <- file.path(tdir, sprintf("gsea_pseudobulk_%s.csv", tag[[pop]]))
  if (!file.exists(f)) return(NULL)
  df <- read.csv(f); df$cell_state <- pop; df
}))
gsea %>%
  transmute(cell_state, signature = pathway_id, NES = round(nes, 3),
            pvalue = signif(pvalue, 3), padj = signif(padj, 3), set_size) %>%
  arrange(signature, factor(cell_state, c("Treg", "Tcon", "CD8"))) %>%
  kable(caption = "Pre-ranked fgsea of mouse WT_heat up/down vs SF-vs-PB pseudobulk (signed moderated t).")
```

**The number that settles it:** `WT_heat_up` NES in **Treg** vs Tcon vs CD8 — is the Treg
NES the largest, and does its FDR clear significance?

```{r money}
up <- gsea %>% filter(grepl("_up$", pathway_id))
if (nrow(up)) {
  treg_nes <- up$nes[up$cell_state == "Treg"]
  cat(sprintf("WT_heat_up NES — Treg: %.2f (padj %.3g) | Tcon: %.2f | CD8: %.2f\n",
              treg_nes,
              up$padj[up$cell_state == "Treg"],
              ifelse(any(up$cell_state=="Tcon"), up$nes[up$cell_state=="Tcon"], NA),
              ifelse(any(up$cell_state=="CD8"),  up$nes[up$cell_state=="CD8"],  NA)))
  cat(sprintf("Treg-preferential? %s\n",
              ifelse(length(treg_nes) && treg_nes >= max(up$nes[up$cell_state!="Treg"], -Inf),
                     "YES — Treg NES is the largest", "NO — not Treg-max (pan-T or elsewhere)")))
}
```

## Secondary (corroborative) — per-cell AUCell score, donor-mean SF vs PB

*The per-cell lens is real AUCell (rank-based, bounded in [0,1], robust to library size and
composition). It stays on its own footing and is never pooled with the pseudobulk NES.*

```{r percell}
eff <- read.csv(here("03_results/master/effect_sizes_treg_arthritis.csv"))
eff %>%
  filter(effect_metric == "percell_auc_smd") %>%
  transmute(cell_state, signature, SMD = round(estimate, 3),
            ci = sprintf("[%.2f, %.2f]", ci_low, ci_high),
            pvalue = signif(pvalue, 3), n_donors) %>%
  kable(caption = "Per-cell WT_heat_up AUCell SF-vs-PB donor-level SMD (secondary tier).")
```

## The effect-size row carried forward to cross-dataset synthesis

```{r effrow}
eff %>%
  filter(evidence_tier == "primary_pseudobulk") %>%
  transmute(dataset, signature, cell_state, contrast, effect_metric,
            estimate = round(estimate, 3), direction, pvalue = signif(pvalue, 3),
            padj = signif(padj, 3), n_donors, n_cells) %>%
  kable(caption = "primary_pseudobulk rows -> master/effect_sizes_treg_arthritis.csv (cross-dataset forest input).")
```

## Figures (rendered by `05_score_signatures_viz.{py,R}`)

*Ordered NES dot plot + per-cell `WT_heat_up` AUCell violins from the `.py` viz. Per-population
running-sum panels (Treg / Tcon / CD8) from the `.R` viz. The violin shows the
`WT_heat_up` arm only, since up and down shift together in SF and AUCell scores each
list unsigned (see the stage README caption). The pseudobulk NES is the primary evidence.*

```{r figs, results='asis'}
ov <- here("03_results/05_scoring/figures/_overview")
for (f in c("wt_heat_nes_forest.png",
            "wt_heat_running_sum_treg.png", "wt_heat_running_sum_tcon.png", "wt_heat_running_sum_cd8.png",
            "score_violins.png")) {
  p <- file.path(ov, f)
  if (file.exists(p)) cat(sprintf("\n![%s](%s)\n\n", f, p))
}
```

### Interactive running-sums — core gene-set distribution across Treg / Tcon / CD8

*The running sums, **interactively**: where the mouse-anchor `WT_heat` core genes fall in
each cell type's SF-vs-PB ranked list. One running-ES line per population (population palette),
with a rug of the core genes below. **Hover a tick** for its symbol, SF-vs-PB signed moderated-t
statistic, and leading-edge status. The drop-down (top-right) switches the **up** / **down**
set. A legend **click** toggles a population's line and rug. Reading it: a Treg-preferential
signal climbs highest in Treg with its core genes packed at the top ranks. A pan-T signal
climbs alike across all three.*

```{r runsum-interactive}
# Interactive running-sum, rendered by the reusable helper sourced in `libs`
# (02_analysis/helpers/runsum_widget.R). Same data logic and same fixed,
# non-overlapping layout (dropdown top-left, legend bottom) as the integration
# report's parallel copy — see the helper header.
build_runsum_widget(here("03_results/05_scoring/tables"))
```

**How to read the per-cell violins vs the ordered NES dot plot.** Per-cell `WT_heat_up`
scores are rank-based **AUCell** values in **[0, 1]**. Absolute level reflects set size and
background detection, so I read the relative SF-vs-PB shift within each population. The dot
plot is a different estimand: fgsea on donor-pseudobulk signed moderated-t rankings. The two
stay separate. Every population shifts SF-high in the same direction. Pseudobulk NES is the
primary evidence.

### The drafted subset — from what's on the map to what I end up with

*The review walks the drafted-subset
design in canonical order — (1) the annotation overview (what is on the map), (2) the
quad marker-context patch (what I draft against), (3) the candidate signatures I draft
for, and (4) the drafted subset I end up with. Every panel is a **design preview /
visualization**, **not a committed cohort and not statistical evidence**. The mouse
anchor heat score is annotation only and is **never a selection gate**. The pseudobulk
NES above remains the evidence the decision rests on. Correlative.*

```{r drafted-flow, results='asis'}
ov <- here("03_results/07_embedding/figures/_overview")
flow <- list(
  list(f = "umap_annotation_treg.png",
       cap = "**(1/4) Annotation overview — what is on the map.** Frozen cell-type annotation (coarse_label: Treg/Tcon/CD8) + tissue (SF/PB) companion. Visualisation only."),
  list(f = "umap_quadmarkers_treg.png",
       cap = "**(2/4) Quad marker context — what I draft against.** %mt + FOXP3 / CTLA4 / IKZF2 (core Treg identity/activation: master lineage TF, suppressive effector, stable-Treg/Helios). Visualisation only."),
  list(f = "umap_signatures_treg.png",
       cap = "**(3/4) Signatures I draft for.** WT_heat (anchor annotation), effector (eTreg) and stress scores landing individually on the sorted T-cell UMAP — the WT_heat overlay is broad (pan-T), not Treg-restricted. Visualisation only."),
  list(f = "umap_drafted_treg.png",
       cap = "**(4/4)** The subset I end up with. Cells the multi-hook OR-union rules would draft (hook_or_union == TRUE, ~33% of the compartment) highlighted vs background, with a facet by which hook drafted them. Design preview, not a committed cohort; anchor heat score never a selection gate.")
)
for (it in flow) {
  p <- file.path(ov, it$f)
  if (file.exists(p)) {
    cat(sprintf("\n%s\n\n![%s](%s)\n\n", it$cap, it$f, p))
  } else {
    cat(sprintf("\n*`%s` not found — run `07_embedding_viz.py` to generate it.*\n\n", it$f))
  }
}
```

## Decision

> **Claim (one sentence):** The mouse 39 °C-derived up arm is **enriched** in JIA
> synovial-fluid T cells relative to paired peripheral blood, and the enrichment is
> **broad across the sorted populations (pan-T), not Treg-preferential**: the running-sum
> concentrates the core gene set at the top of every ranked list, with NES +2.59 in Treg,
> +2.68 in Tcon, and +2.07 in CD8, at FDR 3.2e-14, 8.1e-17 and 3.6e-7. The down arm runs
> at the same sign, reaching significance in Tcon (NES +1.47, FDR 0.026) while staying
> flat in Treg (+0.97) and CD8 (+1.09). Both arms therefore carry direction in Tcon.
>
> **Where the analysis goes next: continue.** This is a positive, reproducible signal
> worth carrying into the cross-dataset comparison, so I proceed to the comparator
> signatures, the effector-Treg correlation, the TCR/activity-inference reads, and the
> other disease datasets. This pseudobulk effect-size row anchors the cross-dataset
> synthesis forest. I read the signal as **consistent with** the mouse stress axis, a
> correlative observation and not evidence of causation, and I state plainly that it is
> **not Treg-specific** — the up-program rises across Treg, Tcon, and CD8 alike. I do not
> overclaim beyond what the synovial-fluid-vs-blood contrast supports.

```{r decisions-echo, echo=TRUE}
str(yaml::read_yaml(here("02_analysis/config/analysis_config.yaml"))$decisions$go_no_go)
```
