---
title: "BREAKPOINT 03 — the project go/no-go: story or null?"
subtitle: "Stage 05_scoring · mouse WT_heat (39 °C Treg) signature in JIA SF-vs-PB"
format:
  html:
    toc: true
    toc-location: left
    df-print: paged
    embed-resources: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.align = "center", dpi = 150, out.width = "95%")
.find_root <- function(sentinel = "02_analysis/config/analysis_config.yaml", start = getwd()) {
  d <- normalizePath(start)
  while (!file.exists(file.path(d, sentinel))) {
    parent <- dirname(d); if (identical(parent, d)) stop("root not found"); d <- parent
  }
  d
}
knitr::opts_knit$set(root.dir = .find_root())
```

```{r libs}
suppressPackageStartupMessages({ library(dplyr); library(knitr) })
here <- function(...) file.path(getwd(), ...)
# Reusable interactive running-sum widget (carved out of the inline chunk below so
# any chunk / any qmd can call it): defines build_runsum_widget().
source("02_analysis/helpers/runsum_widget.R")
```

*Notebook run `r format(Sys.time(), '%Y-%m-%d %H:%M UTC', tz='UTC')`.*

## Where I am — the project pivot

This is the whole-project go/no-go. The
question: does the mouse 39 °C Treg signature show interpretable enrichment in JIA SF Tregs vs
paired blood? What I'm settling here — the one-sentence claim this result licenses (enriched,
with NES/FDR, or a null; both are publishable), whether it's Treg-preferential vs Tcon/CD8 or
just pan-T (a weaker, different claim), which single figure shows it, and what each of go /
pivot / stop actually costs. And if it's null: niche story (keep myeloid/fibroblast scoring
downstream), or "axis invisible in human arthritis" (SAVI stays the positive control)?

> **This is the whole-project go/no-go.** A clean null is a *reported result* the same
> figures carry on equal footing, because the analysis rests on donor-level effect sizes
> with CIs.

## Primary evidence — `WT_heat` NES on donor-pseudobulk SF-vs-PB ranked lists

```{r load-gsea}
tdir <- here("03_results/05_scoring/tables")
tag  <- c(Treg = "treg", Tcon = "tcon", CD8 = "cd8")
gsea <- do.call(rbind, lapply(names(tag), function(pop) {
  f <- file.path(tdir, sprintf("gsea_pseudobulk_%s.csv", tag[[pop]]))
  if (!file.exists(f)) return(NULL)
  df <- read.csv(f); df$cell_state <- pop; df
}))
gsea %>%
  transmute(cell_state, signature = pathway_id, NES = round(nes, 3),
            pvalue = signif(pvalue, 3), padj = signif(padj, 3), set_size) %>%
  arrange(signature, factor(cell_state, c("Treg", "Tcon", "CD8"))) %>%
  kable(caption = "Pre-ranked fgsea of mouse WT_heat up/down vs SF-vs-PB pseudobulk (signed Wald stat).")
```

**Money number:** `WT_heat_up` NES in **Treg** vs Tcon vs CD8 — is the Treg NES the
largest and does its FDR clear significance?

```{r money}
up <- gsea %>% filter(grepl("_up$", pathway_id))
if (nrow(up)) {
  treg_nes <- up$nes[up$cell_state == "Treg"]
  cat(sprintf("WT_heat_up NES — Treg: %.2f (padj %.3g) | Tcon: %.2f | CD8: %.2f\n",
              treg_nes,
              up$padj[up$cell_state == "Treg"],
              ifelse(any(up$cell_state=="Tcon"), up$nes[up$cell_state=="Tcon"], NA),
              ifelse(any(up$cell_state=="CD8"),  up$nes[up$cell_state=="CD8"],  NA)))
  cat(sprintf("Treg-preferential? %s\n",
              ifelse(length(treg_nes) && treg_nes >= max(up$nes[up$cell_state!="Treg"], -Inf),
                     "YES — Treg NES is the largest", "NO — not Treg-max (pan-T or elsewhere)")))
}
```

## Secondary (corroborative) — per-cell donor-mean score, SF vs PB

*Never pooled with the NES: the per-cell score and the pseudobulk NES are different scales.*

```{r percell}
eff <- read.csv(here("03_results/master/effect_sizes_treg_arthritis.csv"))
eff %>%
  filter(effect_metric == "percell_auc_smd") %>%
  transmute(cell_state, signature, SMD = round(estimate, 3),
            ci = sprintf("[%.2f, %.2f]", ci_low, ci_high),
            pvalue = signif(pvalue, 3), n_donors) %>%
  kable(caption = "Per-cell WT_heat_updown SF-vs-PB donor-level SMD (secondary tier).")
```

## The effect-size row that goes to Phase 5

```{r effrow}
eff %>%
  filter(evidence_tier == "primary_pseudobulk") %>%
  transmute(dataset, signature, cell_state, contrast, effect_metric,
            estimate = round(estimate, 3), direction, pvalue = signif(pvalue, 3),
            padj = signif(padj, 3), n_donors, n_cells) %>%
  kable(caption = "primary_pseudobulk rows -> master/effect_sizes_treg_arthritis.csv (Phase-5 forest input).")
```

## Figures (rendered by `05_score_signatures_viz.{py,R}`)

*NES forest + per-cell `WT_heat_up` violins from the `.py` viz; per-population
running-sum panels (Treg / Tcon / CD8) from the `.R` viz. The violin shows the
`WT_heat_up` arm only — up and down shift together in SF, so the combined
up−down metric is uninformative (see the stage README caption); the pseudobulk
NES is the primary evidence.*

```{r figs, results='asis'}
ov <- here("03_results/05_scoring/figures/_overview")
for (f in c("wt_heat_nes_forest.png",
            "wt_heat_running_sum_treg.png", "wt_heat_running_sum_tcon.png", "wt_heat_running_sum_cd8.png",
            "score_violins.png")) {
  p <- file.path(ov, f)
  if (file.exists(p)) cat(sprintf("\n![%s](%s)\n\n", f, p))
}
```

### Interactive running-sums — core gene-set distribution across Treg / Tcon / CD8

*This panel shows the running sums **interactively**: where the mouse-anchor
core gene-set genes fall in each cell type's SF-vs-PB ranked list. One running-ES line per
population (same palette as the forest); the rug ticks below the curves are the mouse
`WT_heat` core signature genes — **hover a tick** for its symbol, its SF-vs-PB signed-Wald
statistic, and whether it sits in the leading edge. Use the drop-down (top-right) to switch
the **up** vs **down** set; **click a population in the legend** to toggle its line + rug.
Reading it: for the up set the Treg curve should climb highest with its core genes packed at
the top ranks (leading edge), and that concentration should look weaker in Tcon / CD8 if the
enrichment is Treg-preferential rather than pan-T.*

```{r runsum-interactive}
# Interactive running-sum, rendered by the reusable helper sourced in `libs`
# (02_analysis/helpers/runsum_widget.R). Same data logic and same fixed,
# non-overlapping layout (dropdown top-left, legend bottom) as the integration
# report's parallel copy — see the helper header.
build_runsum_widget(here("03_results/05_scoring/tables"))
```

**How to read the per-cell violins (secondary tier) vs the forest (primary).** The
per-cell `WT_heat_up` scores sit **negative** even where the biology is positive
(Treg SF > Treg PB) — this is the expected behaviour of the scoring method: `scanpy score_genes` centres each
cell's signature score against a size- and expression-matched **random reference gene
set**, so the absolute level is arbitrary and only the **relative** SF-vs-PB shift within
a population is interpretable (here Treg SF sits clearly above Treg PB). This is a
**different estimand and scale** from the forest. The forest NES is computed at the
backend by **fgsea** (`clusterProfiler::GSEA(by = "fgsea")`) on the **donor-pseudobulk
SF-vs-PB signed-Wald ranked list** — i.e. it asks whether the whole `WT_heat` set
concentrates at the top of the ranked list, then normalises to a NES on its own positive
scale. The two are never pooled: the violins are corroborative
per-cell colour, the pseudobulk NES is the primary evidence and the number the decision
rests on.

### The drafted subset — from what's on the map to what I end up with

*The review walks the drafted-subset
design in canonical order — (1) the annotation overview (what is on the map), (2) the
quad marker-context patch (what I draft against), (3) the candidate signatures I draft
for, and (4) the drafted subset I end up with. Every panel is a **design preview /
visualization**, **not a committed cohort and not statistical evidence**; the mouse
anchor heat score is annotation only and is **never a selection gate**. The pseudobulk
NES above remains the evidence the decision rests on. Correlative.*

```{r drafted-flow, results='asis'}
ov <- here("03_results/07_embedding/figures/_overview")
flow <- list(
  list(f = "umap_annotation_treg.png",
       cap = "**(1/4) Annotation overview — what is on the map.** Frozen cell-type annotation (coarse_label: Treg/Tcon/CD8) + tissue (SF/PB) companion. Visualisation only."),
  list(f = "umap_quadmarkers_treg.png",
       cap = "**(2/4) Quad marker context — what we draft against.** %mt + FOXP3 / CTLA4 / IKZF2 (core Treg identity/activation: master lineage TF, suppressive effector, stable-Treg/Helios). Visualisation only."),
  list(f = "umap_signatures_treg.png",
       cap = "**(3/4) Signatures we draft for.** WT_heat (anchor annotation), effector (eTreg) and stress scores landing individually on the sorted T-cell UMAP — the WT_heat overlay is broad (pan-T), not Treg-restricted. Visualisation only."),
  list(f = "umap_drafted_treg.png",
       cap = "**(4/4) The subset we end up with.** Cells the multi-hook OR-union rules would draft (`hook_or_union == TRUE`, ~33% of the compartment) highlighted vs background, with a facet by which hook drafted them. Design preview, not a committed cohort; anchor heat score never a selection gate.")
)
for (it in flow) {
  p <- file.path(ov, it$f)
  if (file.exists(p)) {
    cat(sprintf("\n%s\n\n![%s](%s)\n\n", it$cap, it$f, p))
  } else {
    cat(sprintf("\n*`%s` not found — run `07_embedding_viz.py` to generate it.*\n\n", it$f))
  }
}
```

## Decision — `decisions.go_no_go`

> **Claim (one sentence):** *(fill at review)* — e.g. "The mouse 39 °C Treg up-program is
> [enriched / not enriched] in JIA SF Tregs vs paired blood (NES = X, FDR = Y), and the
> enrichment is [Treg-preferential / pan-T / absent]."
>
> **Branch:** *(fill at review)* —
> **GO** → proceed to comparators (KO_heat/Interaction), eTreg correlation, TCR/decoupleR,
>   and the other compartments; this row anchors Phase-5 Fig 5A.
> **PIVOT (null in Treg)** → niche story: carry myeloid/fibroblast scoring downstream; or
>   "axis invisible in human arthritis" → SAVI remains the positive control.
> **STOP** → report the null as the primary Phase-1 result.

```{r decisions-echo, echo=TRUE}
str(yaml::read_yaml(here("02_analysis/config/analysis_config.yaml"))$decisions$go_no_go)
```
