---
title: "Is the JIA sorted-T-cell data usable?"
subtitle: "GSE160097 — JIA synovial-fluid vs paired-blood sorted Treg/Tcon/CD8"
format:
  html:
    toc: true
    toc-location: left
    df-print: paged
    embed-resources: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.align = "center", dpi = 150, out.width = "95%")
.find_root <- function(sentinel = "02_analysis/config/analysis_config.yaml", start = getwd()) {
  d <- normalizePath(start)
  while (!file.exists(file.path(d, sentinel))) {
    parent <- dirname(d); if (identical(parent, d)) stop("root not found"); d <- parent
  }
  d
}
knitr::opts_knit$set(root.dir = .find_root())
```

```{r libs}
suppressPackageStartupMessages({ library(dplyr); library(knitr) })
here <- function(...) file.path(getwd(), ...)
```

*Notebook run `r format(Sys.time(), '%Y-%m-%d %H:%M UTC', tz='UTC')`.*

## Where I am

QC has run. This is the static read before I sign off. One question really:
is the data usable? Two things settle it — do enough donors survive with
paired SF+PB Tregs for pseudobulk (it's the donor count, not the cell count, that gives the
forest its power), and do the sorted Treg/Tcon/CD8 separate on their own transcriptomes with no
obvious cross-population bleed? This is a lens, not an analysis. It ends in a recorded
`decisions.qc` call.

## Per-stratum QC yield

```{r load}
tdir <- here("03_results/01_qc/tables")
kept <- read.csv(file.path(tdir, "cells_kept_dropped.csv"))
thr  <- read.csv(file.path(tdir, "qc_thresholds_per_gsm.csv"))
cfg  <- yaml::read_yaml(here("02_analysis/config/analysis_config.yaml"))
floor_cells <- cfg$thresholds$pseudobulk_min_cells
short <- c(CD4_Treg = "Treg", CD4_Tcon = "Tcon", CD8 = "CD8")
kept$pop <- short[kept$population]
kept$tissue_s <- c(synovial_fluid = "SF", peripheral_blood = "PB")[kept$tissue]
```

```{r yield-tab}
kept %>%
  transmute(pop, tissue = tissue_s,
            donor = sub("JIA_patient_", "p", donor),
            n_cells, n_kept, frac_kept = round(frac_kept, 3),
            clears_floor = n_kept >= floor_cells) %>%
  arrange(pop, tissue, donor) %>%
  kable(caption = sprintf("Cells kept per stratum. clears_floor = n_kept >= %d (pseudobulk min).", floor_cells))
```

**Paired-Treg power — the number that matters:**

```{r treg-power}
treg <- kept %>% filter(pop == "Treg", n_kept >= floor_cells)
paired_treg <- treg %>% count(donor, name = "n_tissue") %>% filter(n_tissue == 2)
cat(sprintf("Donors with BOTH SF and PB Treg strata clearing the %d-cell floor: %d / 7\n",
            floor_cells, nrow(paired_treg)))
cat(sprintf("SF-Treg strata retained: %d ; PB-Treg strata retained: %d\n",
            sum(treg$tissue_s == "SF"), sum(treg$tissue_s == "PB")))
```

## Per-sample QC explorer — what is each of the 40 libraries?

*One point per GSM (40 sorted Treg/Tcon/CD8 libraries). x = median genes/cell (log),
y = median mitochondrial %, point size = library size (n_cells), colour = sorted
population. **Hover any point** for its GSM, donor (JIA patient), tissue (synovial
fluid / peripheral blood), sorted population, cell count, kept fraction, and median
counts/genes/%mt, with a kept/EXCLUDED flag. The single hard-excluded library,
**GSM4859852**, sits at far-left low-median-genes and is marked with a black cross and
a pinned label — it is a near-empty library of ambient/empty droplets, a UMI/gene
artifact.*

::: {.content-visible when-format="html"}
```{r qc-explorer, eval=knitr::is_html_output()}
# Interactive per-sample QC explorer, rendered by the reusable helper
# (02_analysis/helpers/qc_sample_explorer.R): aggregates the per-cell QC table to
# one point per GSM. Same separate-regions layout (legend bottom strip) as
# runsum_widget.R so no control overlaps the legend.
source("02_analysis/helpers/qc_sample_explorer.R")
build_qc_sample_explorer(here("03_results/01_qc/tables/qc_metrics_per_cell.csv"))
```
:::

::: {.content-visible when-format="pdf"}
```{r qc-static-pdf, eval=!knitr::is_html_output(), results='asis'}
# Static per-GSM QC violins for the non-interactive (PDF) surface.
p <- here("03_results/01_qc/figures/_overview/qc_violins_per_gsm.png")
if (file.exists(p)) cat(sprintf("\n![qc_violins_per_gsm.png](%s)\n\n", p))
```
:::

## Figures (rendered by `01_qc_filter_viz.py`)

```{r figs, results='asis'}
ov <- here("03_results/01_qc/figures/_overview")
for (f in c("cells_kept_dropped.png", "unsupervised_umap.png")) {
  p <- file.path(ov, f)
  if (file.exists(p)) cat(sprintf("\n![%s](%s)\n\n", f, p))
}
```

## Decision — `decisions.qc`

> **Status: APPROVED.** Yield is high across all strata. One library is hard-excluded.
>
> **GSM4859852** (JIA patient 5, SF CD4 Treg) is a failed near-empty library — ~3,426
> barcodes at ~12–30 UMIs / 3–14 genes each (ambient droplets, not cells). The TCR file is
> a 2.6 KB outlier. Dropped. Disease (JIA) sample. No healthy controls — the contrast is
> SF vs each patient's paired blood. Low criticality: only p5's SF-Treg arm is lost, so
> the Treg SF-vs-blood comparison rests on 6 paired donors. `drop_gsms` keeps it. Data
> usable → freeze the sorted Treg/Tcon/CD8 labels.

```{r decisions-echo, echo=TRUE}
str(yaml::read_yaml(here("02_analysis/config/analysis_config.yaml"))$decisions$qc)
```
