---
title: "Is the JIA sorted-T-cell data usable?"
subtitle: "GSE160097 — JIA synovial-fluid vs paired-blood sorted Treg/Tcon/CD8"
format:
  html:
    toc: true
    toc-location: left
    df-print: paged
    embed-resources: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.align = "center", dpi = 150, out.width = "95%")
.find_root <- function(sentinel = "02_analysis/config/analysis_config.yaml", start = getwd()) {
  d <- normalizePath(start)
  while (!file.exists(file.path(d, sentinel))) {
    parent <- dirname(d); if (identical(parent, d)) stop("root not found"); d <- parent
  }
  d
}
knitr::opts_knit$set(root.dir = .find_root())
```

```{r libs}
suppressPackageStartupMessages({ library(dplyr); library(knitr) })
here <- function(...) file.path(getwd(), ...)
```

*Notebook run `r format(Sys.time(), '%Y-%m-%d %H:%M UTC', tz='UTC')`.*

## Where I am

QC has run; this is the static read before I sign off. One question really:
is the data usable? Two things settle it — do enough donors survive with
paired SF+PB Tregs for pseudobulk (it's the donor count, not the cell count, that gives the
forest its power), and do the sorted Treg/Tcon/CD8 separate on their own transcriptomes with no
obvious cross-population bleed? This is a lens, not an analysis; it ends in a recorded
`decisions.qc` call.

## Per-stratum QC yield

```{r load}
tdir <- here("03_results/01_qc/tables")
kept <- read.csv(file.path(tdir, "cells_kept_dropped.csv"))
thr  <- read.csv(file.path(tdir, "qc_thresholds_per_gsm.csv"))
cfg  <- yaml::read_yaml(here("02_analysis/config/analysis_config.yaml"))
floor_cells <- cfg$thresholds$pseudobulk_min_cells
short <- c(CD4_Treg = "Treg", CD4_Tcon = "Tcon", CD8 = "CD8")
kept$pop <- short[kept$population]
kept$tissue_s <- c(synovial_fluid = "SF", peripheral_blood = "PB")[kept$tissue]
```

```{r yield-tab}
kept %>%
  transmute(pop, tissue = tissue_s,
            donor = sub("JIA_patient_", "p", donor),
            n_cells, n_kept, frac_kept = round(frac_kept, 3),
            clears_floor = n_kept >= floor_cells) %>%
  arrange(pop, tissue, donor) %>%
  kable(caption = sprintf("Cells kept per stratum. clears_floor = n_kept >= %d (pseudobulk min).", floor_cells))
```

**Paired-Treg power — the number that matters:**

```{r treg-power}
treg <- kept %>% filter(pop == "Treg", n_kept >= floor_cells)
paired_treg <- treg %>% count(donor, name = "n_tissue") %>% filter(n_tissue == 2)
cat(sprintf("Donors with BOTH SF and PB Treg strata clearing the %d-cell floor: %d / 7\n",
            floor_cells, nrow(paired_treg)))
cat(sprintf("SF-Treg strata retained: %d ; PB-Treg strata retained: %d\n",
            sum(treg$tissue_s == "SF"), sum(treg$tissue_s == "PB")))
```

## Per-sample QC explorer — what is each of the 40 libraries?

*One point per GSM (40 sorted Treg/Tcon/CD8 libraries). x = median genes/cell (log),
y = median mitochondrial %, point size = library size (n_cells), colour = sorted
population. **Hover any point** for its GSM, donor (JIA patient), tissue (synovial
fluid / peripheral blood), sorted population, cell count, kept fraction, and median
counts/genes/%mt, with a kept/EXCLUDED flag. The single hard-excluded library,
**GSM4859852**, sits at far-left low-median-genes and is marked with a black cross and
a pinned label — it is a near-empty library (ambient/empty droplets), a UMI/gene
outlier rather than a real T-cell profile.*

::: {.content-visible when-format="html"}
```{r qc-explorer, eval=knitr::is_html_output()}
# Interactive per-sample QC explorer, rendered by the reusable helper
# (02_analysis/helpers/qc_sample_explorer.R): aggregates the per-cell QC table to
# one point per GSM. Same separate-regions layout (legend bottom strip) as
# runsum_widget.R so no control overlaps the legend.
source("02_analysis/helpers/qc_sample_explorer.R")
build_qc_sample_explorer(here("03_results/01_qc/tables/qc_metrics_per_cell.csv"))
```
:::

::: {.content-visible when-format="pdf"}
```{r qc-static-pdf, eval=!knitr::is_html_output(), results='asis'}
# Static per-GSM QC violins for the non-interactive (PDF) surface.
p <- here("03_results/01_qc/figures/_overview/qc_violins_per_gsm.png")
if (file.exists(p)) cat(sprintf("\n![qc_violins_per_gsm.png](%s)\n\n", p))
```
:::

## Figures (rendered by `01_qc_filter_viz.py`)

```{r figs, results='asis'}
ov <- here("03_results/01_qc/figures/_overview")
for (f in c("cells_kept_dropped.png", "unsupervised_umap.png")) {
  p <- file.path(ov, f)
  if (file.exists(p)) cat(sprintf("\n![%s](%s)\n\n", f, p))
}
```

## Decision — `decisions.qc`

> **Status: APPROVED.** The QC thresholds are sound. The sorted JIA T-cell libraries
> retain high yield across every stratum, and the only library I hard-exclude is
> **GSM4859852**.
>
> **What GSM4859852 is, and why it is non-critical.** GSM4859852 is the synovial-fluid
> CD4 Treg library of **JIA patient 5** — and it is a **failed, near-empty library**.
> It carries ~3,426 barcodes, but each holds only ~12–30 UMIs and ~3–14 genes, the
> unmistakable signature of ambient/empty droplets rather than intact cells; its paired
> TCR file is a 2.6 KB outlier, far below its siblings. So this is a UMI/gene outlier,
> not a real T-cell profile, and it is right to drop it. It is a **disease (JIA)** sample:
> this cohort has no healthy controls, and the contrast throughout is synovial fluid
> against each patient's own paired peripheral blood. Its criticality is **low** —
> excluding it removes only patient 5's synovial-fluid Treg arm, which is why the Treg
> synovial-fluid-vs-blood comparison downstream rests on **6 paired donors**; the
> enrichment result does not hinge on this one sample. `drop_gsms` already records
> GSM4859852, and it stays there. The interactive explorer above places it exactly where
> its numbers put it — far left, near-empty, marked and labelled apart from the healthy
> mass of the cohort. With that single exclusion, the tissue is usable and I proceed to
> freeze the sorted Treg/Tcon/CD8 labels.

```{r decisions-echo, echo=TRUE}
str(yaml::read_yaml(here("02_analysis/config/analysis_config.yaml"))$decisions$qc)
```
